A Pandemic-Scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-10-08 latex/main.tex

To assess the evidence for these putative novel recombinants

@@

Figure~\ref{fig:recombinant_evidence}A plots the number of mutations required
in the recombinant vs non-recombinant HMM solutions (upper scatter plot)
and the resulting distribution of the number of 
averted mutations (lower histogram)
over all 354 recombination events that pass QC. 
Of these, only 24 
(21 associated with Pango X lineages in Table~\ref{tab:pango_x_lineages},
and 3 additional from Jackon et al in Table~\ref{tab:jackson}) have 
been thoroughly characterised.

Although multiple studies have 
published lists of putatitive novel recombinant 
sequences~\cite{Varabyou2021rapid,Turakhia2022,Li2024CovRecomb} 
these are for GISAID data and not trival to reconcile with 
the Viridian data or the specific recombination events in the ARG.

% CovRecomb:  135,567 putative recombinants across the entirety of 14.5 million
% accessed SARS-CoV-2 genomes
% RIPPLES: In a 1.6 million sample tree from May 2021, we identify 589
% Bolotie: , we identified 225 anomalous SARS-CoV-2 genomes of likely
% recombinant origins out of the first 87,695 genomes to be released
% recombination events, 
% RecombinHunt just looks at X lineages (I think?)

To assess the evidence for these putative novel recombinants

The figure also shows the classification status for each recombinant
according to rebar, a method that classifies sequences
as recombinant or not based on groups of mutations associated with
Pango lineages (STAR Methods). 
Of the 38 recombinants that are strongly supported (at least 10 averted
mutations), 29 are novel. The known recombinants are XA, XC, XS, XBR, 
and 5 from Jackson et al. All but 2 of the 38 are classified as
recombinant by rebar. The majority (20) of these novel strongly supported
recombinants are singletons, with 9 having further descendants in the ARG.
Of the remaining 316 recombinants with less than 10 averted mutations,
only 151 are classified as recombinant by rebar. 
While some are likely to 
be false positives, it is also interesting to note that rebar validation
rate is much lower for recombinants with closely related 
parents (Figure~\ref{fig:recombinant_evidence}B,C) suggesting that 
sc2ts may have a higher sensitivity.

>>>>

@@

Figure~\ref{fig:recombinant_evidence}A plots the number of mutations required
in the recombinant vs non-recombinant HMM solutions (upper scatter plot)
and the resulting distribution of the number of 
averted mutations (lower histogram)
over all 354 recombination events that pass QC. 
Of these, only 24 
(21 associated with Pango X lineages in Table~\ref{tab:pango_x_lineages},
and 3 additional from Jackon et al in Table~\ref{tab:jackson}) have 
been thoroughly characterised.  
To assess the evidence for these putative novel recombination events we 
ran CovRecomb~\cite{Li2024CovRecomb} and rebar on the corresponding sequences 
(STAR Methods). Of the 354 QC-passing sc2ts recombination events, 
90 are classified as 
recombinant by CovRecomb (all but 2 of the 510 QC-failing recombinants 
were classified as non-recombinant).
Rebar classified 203 of the QC-passing sc2ts recombination events, 
as recombinant, with only 10 of those failing QC being classified 
as recombinant.

[WIP]
Of the 38 recombinants that are strongly supported (at least 10 averted
mutations), 29 are novel. The known recombinants are XA, XC, XS, XBR, 
and 5 from Jackson et al. All but 2 of the 38 are classified as
recombinant by rebar. The majority (20) of these novel strongly supported
recombinants are singletons, with 9 having further descendants in the ARG.
Of the remaining 316 recombinants with less than 10 averted mutations,
only 151 are classified as recombinant by rebar. 
While some are likely to 
be false positives, it is also interesting to note that rebar validation
rate is much lower for recombinants with closely related 
parents (Figure~\ref{fig:recombinant_evidence}B,C) suggesting that 
sc2ts may have a higher sensitivity.
